---
title: "Homework 4"
author: "Yingnan Lyu"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
    toc: true
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{The lasso regression vignette}
%\VignetteEncoding{UTF-8}
\usepackage[utf8]{Matrix}
-->

## Homework 4

This vignette answers questions in homework 4 of BIS 557. 

**1.** CASL Number 2 in Exercises 5.8.

```{r}
# Define X and p
X <- matrix(c(1,2,3,
              1,2,5), nrow=3, byrow=TRUE)
X
p <- c(1,1,1)
p

# Show linear Hessian is well-conditioned
solve(t(X) %*% X)

# Check Hessian with logistic variation
var <- diag(p*(1-p))
var
H <- t(X) %*% var %*% X
# solve(H) # This line gives an error due to singularity
```

**2.** CASL Number 4 in Exercises 5.8.

For a genralized linear model with ridge regression penalty, the new log likelihood function is 

$$
l'(y) = l'(y) - \lambda||\beta||_2^2.
$$

Taking the first derivative, and we get the gradient as

$$
\begin{aligned}
\Delta_{\beta}l'(y) = \frac{\delta l'(y)}{\delta \beta_j} 
&= \frac{\delta l(y)}{\delta \beta_j} - 2\lambda\beta\\
&= X^t(y-Ey) - 2\lambda\beta.
\end{aligned}
$$

Taking the second derivative, we get the Hessian matrix as 

$$
\begin{aligned}
H(l') = \frac{\delta^2 l'(y)}{\delta \beta_k \delta \beta_j}
&= \frac{\delta^2 l(y)}{\delta \beta_k \delta \beta_j} - 2\lambda 1_p\\
&= - X_t \cdot \text{diag}(Var(y)) \cdot X - 2\lambda 1_p
\end{aligned}
$$

Using the iteratively reweighted GLM algorithm, we plug in the new Hessian matrix and gradient function and get the updated beta as

$$
\begin{aligned}
\beta^{(k+1)} &= \beta^{(k)} - H^{-1}(l)(\beta^{(k)}\Delta_{\beta}(l)(\beta^{(k)}))\\
&= \beta^{(k)} + [X^t \cdot \text{diag}(Var(y^{(k)})) \cdot X + 2 \lambda 1_p]^{-1} \cdot [X^t(y-Ey^{(k)}) - 2\lambda\beta]\\
&= \beta^{(k)} + V^{-1} \cdot [X^t(y-Ey^{(k)}) - 2\lambda\beta]\\
&= V^{-1}V\beta^{(k)} - 2\lambda V^{-1}\beta + V^{-1} \cdot [X^t(y-Ey^{(k)}) ]\\
&= V^{-1} X^t W X \beta^{(k)} + V^{-1} X^t W \cdot \left(\frac{y-Ey^{(k)}}{diag{Var(y_{(k)})}}\right)\\
&= V^{-1} X^t W \left\{ X \beta^{(k)} +  \frac{y-Ey^{(k)}}{diag{Var(y_{(k)})}} \right\}\\
&=V^{-1} X^t W z
\end{aligned}
$$

where $V = X^t W X + 2 \lambda 1_p$, while W and z are as before.

```{r}
casl_glm_irwls_ridge <-
function(X, y, family, lambda, maxit=25, tol=1e-10)
{
  beta <- rep(0,ncol(X))
  for(j in seq_len(maxit))
    {
    b_old <- beta
    eta <- X %*% beta
    mu <- family$linkinv(eta)
    mu_p <- family$mu.eta(eta)
    z <- eta + (y - mu) / mu_p
    W <- as.numeric(mu_p^2 / family$variance(mu))
    
    XtX <- crossprod(X, diag(W) %*% X)
    V <- XtX + 2*diag(rep(lambda, ncol(X)))
    Xtz <- crossprod(X, W * z)
    beta <- solve(V, Xtz)
    if(sqrt(crossprod(beta - b_old)) < tol) break
    }
  beta
}

# Make data
n <- 1000; p <- 3
beta <- c(0.2, 2, 1)
X <- cbind(1, matrix(rnorm(n * (p- 1)), ncol = p - 1))
mu <- 1 - pcauchy(X %*% beta)
y <- as.numeric(runif(n) > mu)

# Test the function casl_glm_irwls_ridge
beta <- casl_glm_irwls_ridge(X, y,
family = binomial(link = "cauchit"),
lambda = 0.1)
beta
```

**3.** Consider the sparse matrix implementation from class and the sparse add function:

```{r}
sparse_add <- function(a, b) {
  c <- merge(a, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
  c$x <- c$x1 + c$x2
  c[, c("i", "j", "x")]
}

a <- data.frame(i = c(1, 2), j = c(1, 1), x = c(3, 1))
b <- data.frame(i = c(1, 2, 2), j = c(1, 1, 2), x = c(4.4, 1.2, 3))
sparse_add(a, b)
```

        - Implement a `sparse_multiply` function that multiplies two sparse matrices.
        
```{r}
# A function that multiplies two sparse matrices
sparse_multiply <- function(a, b) {
  arow <- max(a$i); acol <- max(a$j)
  brow <- max(b$i); bcol <- max(b$j)
  
  # Create an index of full-rank cols and rows
  index <- data.frame(i=rep(1:max(arow, brow)),
                      j=rep(1:max(acol, bcol), each=max(arow, brow)))
  
  c <- merge(index, a, by = c("i", "j"), all.x = TRUE)
  c <- merge(c, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
 
  for (u in 1:arow) {
    for (v in 1:bcol) {
      c$x[c$i==u & c$j==v] <- sum(c$x1[c$i==u] * c$x2[c$j==v])
    }
  }
  
  c <- c[c$x != 0 & !is.na(c$x),]
  c[, c("i", "j", "x")]
}

a <- data.frame(i = c(1, 2), j = c(1, 1), x = c(3, 1))
b <- data.frame(i = c(1, 2, 2), j = c(1, 1, 2), x = c(4.4, 1.2, 3))
sparse_multiply(a, b)
```

        - Create a new class `sparse.matrix` that has add `+`, multiply `%*%`, and transpose `t()` methods.
        
```{r}
# Define a function to generate sparse matrix
sparse.matrix <- function(i, j, x, dims=NULL) {
  
  sm <- data.frame(i=i, j=j, x=x)
  
  # If no dimensions are given, assume the largest coordinates are the dimensions.
  if (is.null(dims)) {
    dims <- c(max(i), max(j))
  } 
  
  r <- dims[1]; c <- dims[2]
  index <- data.frame(i=rep(1:r), j=rep(1:c, each=r))
  sm <- merge(index, sm, by = c("i", "j"), all.x = TRUE, sort=FALSE)
  sm <- sm[sm$x != 0 & !is.na(sm$x),]
  
  attr(sm$i, "label") <- r
  attr(sm$j, "label") <- c
  class(sm) <- c("sparse.matrix", class(sm))
  sm
}



# Define function + to add sparse matrix
`+.sparse.matrix` <- function(a,b) {
  # Check whether the dimensions of a and b agree
  if (attr(a$i, "label") != attr(b$i, "label") ||
      attr(a$j, "label") != attr(b$j, "label")) {
    stop('The dimensions of the two input matrix do not match')
  }
  
  c <- merge(a, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
  c$x <- c$x1 + c$x2
  c <- c[, c("i", "j", "x")]
  
  #Make a sparse matrix
  sm <- data.frame(i=c$i, j=c$j, x=c$x)
  dims <- c(attr(a$i, "label"), attr(a$j, "label"))
  
  r <- dims[1]; c <- dims[2]
  index <- data.frame(i=rep(1:r), j=rep(1:c, each=r))
  sm <- merge(index, sm, by = c("i", "j"), all.x = TRUE, sort=FALSE)
  sm <- sm[sm$x != 0 & !is.na(sm$x),]
  
  attr(sm$i, "label") <- r
  attr(sm$j, "label") <- c
  class(sm) <- c("sparse.matrix", class(sm))
  sm
}



# Define function t to transpose sparse matrix 
t.sparse.matrix <- function(x) {
  # swap i and j index
  ta <- x
  names(ta) <- c('j', 'i', 'x')
  ta <- ta[,c('i', 'j', 'x')]
  
  r <- attr(ta$i, "label"); c <- attr(ta$j, "label")
  index <- data.frame(i=rep(1:r), j=rep(1:c, each=r))
  ta <- merge(index, ta, by = c("i", "j"), all.x = TRUE, sort=FALSE)
  ta <- ta[ta$x != 0 & !is.na(ta$x),]
  
  #Make a sparse matrix
  sm <- data.frame(i=ta$i, j=ta$j, x=ta$x)
  dims <- c(r, c)
  
  index <- data.frame(i=rep(1:r), j=rep(1:c, each=r))
  sm <- merge(index, sm, by = c("i", "j"), all.x = TRUE, sort=FALSE)
  sm <- sm[sm$x != 0 & !is.na(sm$x),]
  
  attr(sm$i, "label") <- r
  attr(sm$j, "label") <- c
  class(sm) <- c("sparse.matrix", class(sm))
  sm
}



# Define a function %*% to multiply sparse matrix
`%*%.default` = .Primitive("%*%")  

`%*%` = function(x,...){ 
  UseMethod("%*%",x)
}

`%*%` <- function(x, y) {
  UseMethod("%*%", x)
}

`%*%.sparse.matrix` <- function(a,b) {
  #UseMethod("%*%", a)
  # Check the type of b
  if (!inherits(b, "sparse.matrix")) {
    stop("y argument is not a sparse matrix type.")
  }
  # Check whether the dimensions of a and b agree
  if (attr(a$j, "label") != attr(b$i, "label")) {
    stop('The dimensions of the two input matrix do not match')
  }
  
  if (attr(a$i, "label") == 0 || attr(a$j, "label") == 0 ||
      attr(b$i, "label") == 0 || attr(b$j, "label") == 0) {
    arow <- max(a$i); acol <- max(a$j)
    brow <- max(b$i); bcol <- max(b$j)
  } else {
    arow <- attr(a$i, "label"); acol <- attr(a$j, "label")
    brow <- attr(b$i, "label"); bcol <- attr(b$j, "label")
      }
  
  # Create an index of full-rank cols and rows
  index <- data.frame(i=rep(1:max(arow, brow)),
                      j=rep(1:max(acol, bcol), each=max(arow, brow)))
  
  c <- merge(index, a, by = c("i", "j"), all.x = TRUE)
  c <- merge(c, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
 
  for (u in 1:arow) {
    for (v in 1:bcol) {
      c$x[c$i==u & c$j==v] <- sum(c$x1[c$i==u] * c$x2[c$j==v])
    }
  }
  
  c <- c[c$x != 0 & !is.na(c$x),]
  c <- c[, c("i", "j", "x")]
  
  #Make a sparse matrix
  sm <- data.frame(i=c$i, j=c$j, x=c$x)
  dims <- c(attr(a$i, "label"), attr(b$j, "label"))

  r <- dims[1]; c <- dims[2]
  index <- data.frame(i=rep(1:r), j=rep(1:c, each=r))
  sm <- merge(index, sm, by = c("i", "j"), all.x = TRUE, sort=FALSE)
  sm <- sm[sm$x != 0 & !is.na(sm$x),]
  
  attr(sm$i, "label") <- r
  attr(sm$j, "label") <- c
  class(sm) <- c("sparse.matrix", class(sm))
  sm
}
```

        - Add test-sparse-matrix.r to the testthat directory of your bis557 package to show it works.

